---
title: "Compare SSDM Files"
output:
  html_document: default
Author: Casey Breen
---

### Library Packages

```{r}
library(data.table)
library(tidyverse)
```

### Read SSDM Files and append together

```{r, eval=FALSE}
SSDM1 <- read_fwf("/data/josh//CenSoc/ssdm/ssdm1",
                          fwf_widths(c(1,9, 20, 4, 15, 15, 1, 8, 8),
                                     col_names = c("mode", "ssn", "lname",
                                                   "name_suffix", "fname", "mname",
                                                   "vorpcode", "dod", "dob")))
SSDM2 <- read_fwf("/data/josh//CenSoc/ssdm/ssdm2",
                          fwf_widths(c(1,9, 20, 4, 15, 15, 1, 8, 8),
                                     col_names = c("mode", "ssn", "lname",
                                                   "name_suffix", "fname", "mname",
                                                   "vorpcode", "dod", "dob"))) 
SSDM3 <- read_fwf("/data/josh//CenSoc/ssdm/ssdm2",
                          fwf_widths(c(1,9, 20, 4, 15, 15, 1, 8, 8),
                                     col_names = c("mode", "ssn", "lname",
                                                   "name_suffix", "fname", "mname",
                                                   "vorpcode", "dod", "dob"))) 

SSDM_full <- rbind(SSDM1, SSDM2, SSDM3)
rm(SSDM1, SSDM2, SSDM3)
```

### Read SSDM NBER Files and append together

```{r, eval=FALSE}
SSDM1_nber <- fread("/data/josh/CenSoc/ssdm/nber_versions/ssdm1_nber.csv")
SSDM2_nber <- fread("/data/josh/CenSoc/ssdm/nber_versions/ssdm1_nber.csv")
SSDM3_nber <- fread("/data/josh/CenSoc/ssdm/nber_versions/ssdm1_nber.csv")

SSDM_full_nber <- rbind(SSDM1_nber, SSDM2_nber, SSDM3_nber)
rm(SSDM1_nber, SSDM2_nber, SSDM3_nber)
```

### save out .rda files

```{r, eval=FALSE}
save(SSDM_full, file = "/90days/casey/ssdm_full.rda")
save(SSDM_full_nber, file = "/90days/casey/ssdm_full_nber.rda")
```

# Load SSDM Datasets
```{r, eval=FALSE}
SSDM_full <- as.data.table(load("/90days/casey/ssdm_full.rda"))
SSDM_full_nber <- as.data.table(load("/90days/casey/ssdm_full_nber.rda"))

SSDM_full <- as.data.table(SSDM_full)
SSDM_full[,"dyear" := as.numeric(substr(dod, 5, 9))]

substrRight <- function(x, n){
  substr(x, nchar(x)-n+1, nchar(x))
}

SSDM_full_nber[,"dyear" := as.numeric(substrRight(dod, 4))]
```

### Tabulate by deaths, combine death tabulations, calculate measures for graphing

```{r, eval=FALSE}
SSDM_table <- SSDM_full[, .N, by = list(dyear)]
SSDM_table_nber <- SSDM_full_nber[, .N, by = list(dyear)]

```

```{r}
load("/90days/casey/ssdm_full_nber_table.rda")
load("/90days/casey/ssdm_full_table.rda")


ssdm_table_combined <- SSDM_table_nber %>%
  full_join(SSDM_table, by = "dyear") %>%
  rename(nber_count = N.x, count = N.y) %>%
  mutate(count_diff = nber_count - count) %>%
  mutate(percent_diff = (nber_count/count)*100)
  
```

### Compare Death Tabulations

```{r}
ggplot(ssdm_table_combined)+ 
  geom_point(aes(x = dyear, y = nber_count, colour = "blue")) +
  geom_point(aes(x = dyear, y = count, colour = "red"))         
```
```{r}
ssdm_table_combined_2005 <- as.tibble(ssdm_table_combined %>%
  filter(dyear > 2003 & dyear < 2011))

```

|dyear| nber_count|   count| count_diff| percent_diff|
|-----:|----------:|-------:|----------:|------------:|
|  2004|    2192040| 2213874|     -21834|     99.01377|
|  2005|    2237958| 2183019|      54939|    102.51665|
|  2006|    2224995| 2097920|     127075|    106.05719|
|  2007|    2243583| 1855977|     387606|    120.88420|
|  2008|    2285637| 1638857|     646780|    139.46531|
|  2009|    2257212| 1525359|     731853|    147.97907|
|  2010|     359622| 1424464|   -1064842|     25.24613|


## Difference in total count of deaths (NBER_SSDM - SSDM) 

```{r}
ggplot(ssdm_table_combined)+ 
  geom_point(aes(x = dyear, y = count_diff)) 
```

## Percent difference in deaths counted (NBER_SSDM/SSDM) 

```{r}
ggplot(ssdm_table_combined)+ 
  geom_point(aes(x = dyear, y = percent_diff)) 
```


## Percent difference in deaths
```{r}
ssdm_table_filtered <- ssdm_table_combined %>%
  filter(dyear > 1980 & dyear < 2010) 

ggplot(ssdm_table_filtered)+ 
  geom_point(aes(x = dyear, y = percent_diff))
```




