---
title: "Create Censoc Datasets"
output: html_notebook
---

### Set up  

- Create a new local project
    - New project from version control
    - Clone censoc from Github:  https://github.com/MJAlexander/censoc
    - Add removed censoc functions to bottom of NAMESPACE (load_socsec_deaths, load_census, create_national_censoc)
- Update path names
    
### Load Packages

```{r}
library(censoc)
library(tidyverse)
library(data.table)
```

### State Match

```{r, eval = FALSE}
# keep a log

update_file <- "//home/ipums/casey-ipums/censoc/make_censoc/progress.txt"

matched_data_folder <- "//home/ipums/casey-ipums/censoc/make_censoc/states/"

# load in the socsec files

socsec_files = c("/home/ipums/josh-ipums/progs/ssdm/ssdm1",
                     "/home/ipums/josh-ipums/progs/ssdm/ssdm2",
                     "/home/ipums/josh-ipums/progs/ssdm/ssdm3")

socsec <- load_socsec_deaths(socsec_files)
sink(update_file)
cat("Socsec has been read in. \n")

# get the list of state census files

state_file_list <- list.files("/home/ipums/casey-ipums/IPUMS/1940/CSV/")
state_fips <- as.numeric(gsub("\\D", "", state_file_list))
state_fips <- state_fips[!is.na(state_fips)]
state_file_list <- paste0("/home/ipums/casey-ipums/IPUMS/1940/CSV/", state_file_list[grepl("state", state_file_list)])



cols_to_keep <- c("URBAN", "OWNERSHP", "RENT", "RACE", "MARST", "BPL", "HISPAN",
                  "EMPSTAT", "OCCSCORE", "INCWAGE","INCNONWG",
                  "EDUC",
                  "MBPL", "FBPL", "OCC")

# loop over each state file, do the match, and save the matched output

for(i in 1:length(state_file_list)){
  cat(paste("Reading in census file", i, "\n"))
  census <- try(load_census(census_file = state_file_list[i], cols_to_keep = cols_to_keep))
  if(class(census)=="try-error"){
    cat(paste("File", i, "is empty \n"))
  }
  else{
    cat(paste("Matching census file", i, "\n"))
    censoc <- create_censoc(census, socsec)
    cat(paste("Writing file", i, "\n"))
    write_csv(censoc, paste0(matched_data_folder, state_fips[i], "_matched.csv"))
  }
}

rm(socsec)

sink()

```

- Several warnings messages (1 per state): the condition has length > 1 and only the first element will be used

### National match
```{r, eval = FALSE}
# state censoc files
files  <- paste0("~/censoc/make_censoc/states//",list.files(path = "~/censoc/make_censoc/states//", pattern = '\\.csv') )

# national censoc
all_censoc_unique <- create_national_censoc(files)

# for BFDW2018
cols_to_keep <- c("URBAN", "OWNERSHP", "RENT", "RACE", "MARST", "BPL", "HISPAN",
                  "EMPSTAT", "OCCSCORE", "INCWAGE","INCNONWG",
                  "EDUC",
                  "MBPL", "FBPL", "OCC")
bfdw <- all_censoc_unique[, c("SERIAL40", "NUMPREC40", "PERNUM", "STATEFIP",  "census_age.x", "byear", "dyear", "bmonth", "dmonth", cols_to_keep)]
bfdw <- bfdw %>% rename(census_age = census_age.x)
write_csv(bfdw, path  = "~/censoc/make_censoc/censoc_bfdw.csv")

# save dataset with less variables
dr <- all_censoc_unique[, c("SERIAL40", "NUMPREC40", "PERNUM", "STATEFIP",  "census_age.x", "byear", "dyear", "bmonth", "dmonth")]
write_csv(dr, "~/censoc/make_censoc/national_matched_states_red.csv")

```

## Questions
- national matched red states? 
- dday missing and bday missing? Why are we giving bday missing and dday missing instead of actual dates? Confidentiality?

```{r, eval = FALSE}
colsToKeep <- c("SERIAL40", "NUMPREC40", "PERNUM", "STATEFIP",  "census_age", "byear", "dyear", "bmonth", "dmonth")

censoc <- fread("/90days/casey/censoc_id_date_v1.csv", header=TRUE, select=colsToKeep)
```

Here can compare means to check how similar the datasets are (both contain 7654451 matches)

```{r, include=TRUE}

censoc <- as.data.table(censoc)
d <- censoc[,lapply(.SD, mean)]

censoc_casey <- as.data.table(dr)
d2 <- censoc_casey[,lapply(.SD, mean)] %>% 
  rename(census_age = census_age.x)

means <- bind_rows(d, d2)
means <- rbind(means, means[1, ] - means[2, ])

```

## Compare Means for both Datasets


```{r}
means
```

