---
title: "Accessing Censoc with full IPUMS vars"
output: html_notebook
author: Casey Breen
---


## Introduction

In this notebook, I explore the data quality of the CenSoc with full IPUMS variables. 

We hope to: 

(1) Explore and Identify and erroneous data in the full Censoc
    
(2) Decide whether these cases should be removed from the dataset

    - Women being included in the dataset
    - Strange names, characters, etc. 


## Read in full CenSoc Data
```{r}
library(data.table)
```


```{r}
dt <- fread("~josh-ipums/censoc/full_censoc_censoc.csv")
head(dt)
```
We get 7469522 rows. 


Let's check that there are no women in the dataset: 

```{r}
dt[SEX != 1, .(NAMEFRST, SEX, RELSTR)]
```


RELSTR contains the alphabetic string describing an individual's relationship to the head of household or householder (from IPUMS). Daughter, Maid and Wife might be as issues Let's check these out: 

```{r}
dt[RELSTR %in%  c("'Daughter'", "'Maid'", "'Wife'") & SEX ==1, .(NAMEFRST, SEX, RELSTR)]
```

Are there any non-alpha characters (besides spaces)? Nope! 

```{r}
dt[!grepl("[A-z]", NAMEFRST), .(NAMEFRST)]
```

Are there any short names? I'm checking for the actual first or last name containing fewer than 3 characters. Note that `` take up two characters. Looks like there's around ~5,000.  

```{r}
dt[(nchar(NAMEFRST) < 5) | (nchar(NAMELAST) < 5), .(NAMEFRST, NAMELAST)]
```

```{r}
dt_short_names <- dt[(nchar(NAMEFRST) < 5) | (nchar(NAMELAST) < 5), .(NAMEFRST, NAMELAST)]

table <- dt_short_names[, .N ,by = NAMEFRST]

table[order(-N),]
```



How about where the first and the last names are short? Nope! 

```{r}
dt[(nchar(NAMEFRST) < 5) & (nchar(NAMELAST) < 5), .(NAMEFRST, NAMELAST)]
```

Let's take the top names women's names (that were also not top men's names) from the SSA: https://www.ssa.gov/oact/babynames/decades/names1900s.html for 1900, 1910, 1920, 1930, 1940.

```{r}
popular_names <- read_tsv("/home/ipums/casey-ipums/censoc/casey/top_names_1900_1940", col_names = c("rank", "men", "freq_men", "women", "freq_women")) %>%
  select(men, women)

womens_names <- paste0("'", unique(popular_names$women[!popular_names$women %in% popular_names$men]), "'")

head(womens_names)
```

How many men have popular womens names? 40,464. 

```{r}
dt[NAMEFRST %in% womens_names,.(NAMEFRST, SEX, RELSTR)]
```

How many men have both a women's name and a female relationship string? Maybe these cases we should drop! 


```{r}
dt[NAMEFRST %in% womens_names & SEX == 1 &RELSTR %in% c("'Wife'", "'Daughter'", "'Maid'"),.(NAMEFRST, SEX, RELSTR)]
```










