---
title: "Truncated Gompertz mortality estimation by cohort and by race"
output:
  html_document: default
  html_notebook: default
---

As the observed deaths in the CenSoc dataset are truncated to be within a certain age range which varies across cohorts, any comparison across cohorts of mortality will be unfair. Therefore, the dataset must adjust for the truncation. Fortunately, a Bayesian truncated Gompertz model can be used for this dataset. 

The purpose of this vignette to present useful tools from the `censoc` R package and to show an application of the Bayesian truncated Gompertz model. For this case, the example will be an estimation of life expectancy at age 65 (i.e. the average years left to live at age 65) by cohort and race, for cohorts 1895-1905 using the CenSoc dataset.

Please review the following links for more information on the [CenSoc project](https://censoc.demog.berkeley.edu/), [dataset details](https://censoc.demog.berkeley.edu/articles/match_document.html) and the [Bayesian truncated Gompertz model](https://censoc.demog.berkeley.edu/articles/censoc_chapter.pdf).

##Short example code:
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(rjags)
library(R2jags)
library(censoc)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
censoc_file_path <- "/data/josh/CenSoc/censoc_bfdw.csv"
df <- read_csv(file = censoc_file_path, progress = FALSE)
```

####Variables and death counts
After loading the necessary packages and data, we create the variables of the age of death and race categories using functions from the `censoc` package. 

```{r}
# Variables
df <- add_age_of_death_column(df) #age of death
df<-add_race_recode_column(df) #covariate of interest: race

df$race_cat <- NA #Reducing race into fewer categories in a new variable
df$race_cat[df$race_recode=="White"] <- 1
df$race_cat[df$race_recode=="Black"] <- 2
df$race_cat[df$race_recode=="American Indian/Alaska Native" | df$race_recode=="Chinese"
            | df$race_recode=="Filipino" 
            | df$race_recode=="Japanese"  ] <- 3


# Tabulating the deaths:
tab_df <- tabulate_deaths(df, race_cat)

tab_df$race_label <- NA # defining the labels for each race category
tab_df$race_label[tab_df$race_cat==1] <- "White"
tab_df$race_label[tab_df$race_cat==2] <- "Black"
tab_df$race_label[tab_df$race_cat==3] <- "Other"

cohorts <- 1895:1915#1905
tab_df <- filter(tab_df, byear %in% cohorts) # data only contains these cohorts

```

####Running the model
```{r}
min_age <- 65
meta_data <- get_meta_data(tab_df, min_age)
group_labels <- meta_data$group_labels

# Model
mod <- truncated_gompertz_model(tab_df, min_age, number_chains = 4, number_iterations = 10000, run_in_parallel = TRUE)
```

```{r, message=FALSE}
# Life expectancy (ex) and hazard rate (hx) estimates
res <- get_estimated_hx_ex(mod, ncohorts, ngroups, group_labels, nages, alpha_level = 0.05)

ex_df <- res$ex_df # extract life expectancy dataframe

```
#### Results: life expectancy at age 65 by race and cohort

```{r, message=FALSE}
group_title <- "Race"

ex_age <- 65# Age of life expectancy to be plotted 

#Plot
plot_ex(ex_df, ex_age, group_title)
```

###Important comments
* For this example, we use the cohort of people born between 1895 and 1905. However, this cohort may be modified to include any years between 1865 to 1940.

* Computationally, the minimum age can be as low as 0. However, in terms of the Gompertz mortality model, we are interested in adults. So, `min_age` should be greater than `insert AGE`. 

* The maximum age is not set by the user, it its fixed at 110. However, setting `ex_age` at 110 will yield a graph with no useful results. 

* The `tabulate_deaths` function removes any observations whose year of death falls outside in the period 1975-2005. It also restricts the age of death to be in between 65 and 110. 

* For the `truncated_gompertz_model` function to work the `tab_df` must follow the ordering used above. In particular, the third column must be a number referring to the group level. The fifth column must be a label for each group.

* For the model to converge, the value of "Rhat" must be 1.1 or lower. If your model did not converge, try running it for more iterations. 
__NOTE: Model run time can be quite long.__
  


 