---
title: "SS-5 Coverage Notes"
output: html_notebook
---

## Introduction

In this notebook, I look at the completeness of the NUMIDEATH files we received from the National Archives.

Our goals are:

1. To see how counts in NUMIDEATH compare to HMD death counts for entire US population.

2. To try to figure out the reason for the differences, perhaps by looking at coverage rates by state.

## Reading NUMIDEATH files and data cleaning

```{r}
library(data.table)
dt <- fread("numdeath_all.csv") ## this is a combined file that I created with only a few variables
```
We get 49 million rows. 

```{r}
head(dt)
```
And we can see that we have only SSN, first name, last name, dob, dod.


We noticed that some of the dates are missing a zero before the month.

```{r}
## check dod and dob strings for length
table(nchar(dt$dod))
```
```{r}
table(nchar(dt$dob))
```
We could fix these, but it takes time. Just drop for now.
```{r}
dt <- dt[nchar(dob) != 7]
```
We create separate variables for year, month and day.
```{r}
## step 1, get dates
dt[, dyear := as.numeric(substr(dod, 5, 8))]
dt[, byear := as.numeric(substr(dob, 5, 8))]
dt[, bmonth := as.numeric(substr(dob, 1,2))]
dt[, dmonth := as.numeric(substr(dod, 1,2))]
dt[, bday := as.numeric(substr(dob, 3,4))]
dt[, dday := as.numeric(substr(dod, 3,4))]
```
For the birth and death records missing "day", we impute.
```{r}
## fix dday and bday if "XX" or "0" by imputing 1st of month
dt[bday == 0 | is.na(bday), .N]
dt[bday == 0 | is.na(bday), bday := 1]
dt[dday == 0 | is.na(dday), .N]
dt[dday == 0 | is.na(dday), dday := 1]
```
We calculate age at death with an accuracy of about a month.
```{r}
## decent approx of age at death
dt[, age.at.death := (dyear + dmonth/12) - (byear + bmonth/12)]
```
Drop people who die before they're born
```{r}
dt[age.at.death < 0, .N] ## only 124 cases
dt <- dt[age.at.death >= 0,]
## visualize
hist(dt$age.at.death)
## so it looks like we have a few very old people >> 110
```

## Lexis table of deaths by year and age

```{r}
dt[, years.at.death := floor(age.at.death)]
Dxt.tab <- dt[, table(years.at.death, dyear)]
```

## visualize

```{r}
filled.contour(x = as.numeric(rownames(t(Dxt.tab))),
               y = as.numeric(colnames(t(Dxt.tab))),
               log(t(Dxt.tab)))
## let's look only at post 1960 and age under 110

my.Dxt.tab <- Dxt.tab[paste(50:110), paste(1980:2007)]
filled.contour(x = as.numeric(rownames(t(my.Dxt.tab))),
               y = as.numeric(colnames(t(my.Dxt.tab))),
               log(t(my.Dxt.tab)))

my.Dxt.tab <- Dxt.tab[paste(65:96), paste(1985:2007)]
filled.contour(x = as.numeric(rownames(t(my.Dxt.tab))),
               y = as.numeric(colnames(t(my.Dxt.tab))),
               log(t(my.Dxt.tab)))
filled.contour(x = as.numeric(rownames(t(my.Dxt.tab))),
               y = as.numeric(colnames(t(my.Dxt.tab))),
               t(my.Dxt.tab))
```
## What is start date where we have fuller reporting?

```{r}
tt <- table(dt[dyear > 1950]$dyear)
plot(tt)
```


So it looks like 1988 is the year when reporting increases. Perhaps we should be concerned that there's a decline after 2002 or so.

NCHS reports 2.4 million deaths in 2000. Here we see

```{r}
tt["2000"]
```
2.2 million. So we're doing pretty well. Not perfect but pretty well.

## Comparison with HMD

```{r}
hmd <- fread("/data/josh/CenSoc/hmd/hmd_statistics/deaths/Deaths_1x1/USA.Deaths_1x1.txt")
hmd[Age == "110+", Age := "110"]


## now create an Dxt matrix
hmd.Dxt <- xtabs(Total ~ Age + Year, hmd)
```

```{r, fig.asp = 1.5}
## now let's compare
Dxt.ss5 <- Dxt.tab[paste(50:110), paste(1985:2007)] 
Dxt.hmd <-   hmd.Dxt[paste(50:110), paste(1985:2007)]
ratio.mat <- t(Dxt.ss5/Dxt.hmd)
my.breaks <- c(0, .8, seq(.9, .99, .01))
image(x = as.numeric(rownames(ratio.mat)),
      y = as.numeric(colnames(ratio.mat)),
      z = ratio.mat, 
      breaks = my.breaks, 
      col = heat.colors(length(my.breaks)-1),
      xlab = "year",
      ylab = "age")
contour(x = as.numeric(rownames(ratio.mat)),
      y = as.numeric(colnames(ratio.mat)),
      z = ratio.mat, 
levels = c(seq(0, .9, .1), seq(.91, .99, .01)),
add = TRUE)
abline(v = c(1988, 2004), lwd = 4, col = "blue")
abline(h = 65, lwd = 4, col = "blue")

```

## let's look by  cohort

```{r}
## restructure the data so it has cohort
hmd.dt <- as.data.table(Dxt.hmd)
ss5.dt <- as.data.table(Dxt.ss5)
names(ss5.dt) <- c("Age", "Year", "N")
hmd.dt[, x := as.numeric(Age)]
hmd.dt[, year := as.numeric(Year)]
ss5.dt[, x := as.numeric(Age)]
ss5.dt[, year := as.numeric(Year)]
hmd.dt[, cohort := year - x]
ss5.dt[, cohort := year - x]

```

```{r}
par(mfrow = c(2,3))
cohort.vec <- c(1890, 1900, 1910, 1920, 1930)
for (i in 1:length(cohort.vec))
{
  this.coh <- cohort.vec[i]
hmd.dt[cohort == this.coh, plot(x, N, ylim = c(100, max(N)),
##                                log = "y",
                               type = "l", col = "blue")]
ss5.dt[cohort == this.coh, lines(x, N, lty = 2, col = "red")]
title(paste("Cohort of", this.coh))
}

```
So we can see pretty clearly that we start to be pretty good about 1988 through 2002, or so.

## Concluding thoughts

* Coverage is remarkably complete after 1988. So it's not that the "state death" issue is happening all along. 

* The possibility of reweighting for places where we have undercount is worth thinking about -- particularly how to test the assumption that the chance of being counted is independent of characteristics other than age and cohort (e.g., race, educ, state)

* Next steps include linking to SSDM or to CENSOC to see who is left out.




